\( \tool \): JavaScript IR-based Semantics Extraction Toolchain

### 2020-04-13 ase20/compiler.tex
Revised Section 4

\section{Algorithm Compiler}\label{sec:compiler}
\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{img/algo-compiler.pdf}
  \caption{overall structure of {\sf algorithm compiler}}
  \label{fig:algo-compiler}
\end{figure}

@@

Moreover, {\sf Tokenizer} flattens a structured step to a single token
list to handle multi-step statements easily.  Some statements in
abstract algorithms consist of multiple steps.  For example, the
\( \code{if}\!-\!\code{then}\!-\!\code{else} \) statement often consists of
two steps: one for the \( \code{then} \)-branch and another
for the \( \code{else} \)-branch.  To treat them as a linear structure,
we introduce three special tokens to break down structured algorithms:
\( \tend \) denotes the end of a single step, and \( \tin \) and
\( \tout \) denote the start and the end of nested steps, respectively.
For example, the following left abstract algorithm is tokenized to the right
token list.
\[
  \begin{array}{lcl}
    \code{1. A}\\
    \code{2. B} &\Longrightarrow& \code{A} \tend \code{B} \tin \code{C} \tend \tout \tend\\
    \code{\ \ a. C}\\
  \end{array}
\]

@@

\subsection{Token List Parser}
The token list parser is defined with \textit{parsing rules}. Each parsing rule
is a basic parsing rule or a composition of other parsing rules. The composition
\code{A | B} of two parsing rules \code{A} and \code{B} parses the input using
both rules and collect the longest matched results.  If both rules are failed or
matched the same length of the input, the composition is failed.

We provide two kinds of basic parsing rules: \textit{tag-based rules} and
\textit{content-based rules}.  A tag-based rule just checks whether the next
token has a given tag.  For example, the tag-based parser \( \code{varT} \) and
\( \code{codeT} \) check whether the next token has the tag \( \code{<var>} \)
and \( \code{<code>} \), respectively.  A content-based parser checks whether
the next token is a text token and its content passes a given condition.  For
example, the string literal \( \code{"Perform"} \) denotes a content-based
parser that checks whether the next token is a text token with the content \(
\code{Perform} \).  We also define two content-based parsers \( \code{word} \)
and \( \code{number} \) that check whether the content of the next token
consists of only alphabets or numbers, respectively.  In addition, we provide
several helper functions in parsing rules such as optional rule \code{A?},
positive(negative) predicate \code{+A}(\code{-A}).  For instance, the helper
function \( \code{repsep(A, B)} \) generates a new parsing rule that denotes
zero or more repetition of the parsing rule \( \code{A} \) using another parsing
rule \( \code{B} \) as a separator.

@@

\subsection{Token AST Converter}
\textit{Conversion rules} describe how to generate an \( \ires \) function for a
given token AST.  Each conversion rule is defined with its corresponding parsing
rule.  For basic parsing rules, their conversion rules always return the string
values of the contents in parsed tokens.  For example, the following conversion
rules are the omitted parts in the previous example for the step 5 of
\textbf{Evaluation}:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ e ~ _ => IExpr(e) }
// expressions
val Expr =
  // codes      // false literal
  ... ^^ EStr | ... ^^ { _ => EBool(false) }            |
  // variables  // additions
  ... ^^ EId  | ... ^^ { case x ~ _ ~ y => EAdd(x, y) } |
  // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ECall(x, y) }
)
\end{lstlisting}
The conversion rule of the \code{Stmt} compile rule uses only the second sub
tree and constructs an \code{IExpr} \( \ires \) instruction.  For the second
sub-tree, the conversion rule of the fourth \code{Expr} compile rule is
applied. It constructs \code{ECall} \( \ires \) expression with the string value
of the first sub-tree and the sequence of the expressions of the third sub-tree.
In this way, the step 5 of \textbf{Evaluation} is converted to the following \(
\ires \) instruction whose beautified form is the sixth line of the function
body in Figure~\ref{fig:array-literal-eval}.
\begin{lstlisting}[style=ires]
IExpr(ECall(EId("Set"), List(
  EId(array), EStr("length"), ECall(EId(ToUint32), List(
    EAdd(EId("padding"), EId("len")))), EBool(false))))
\end{lstlisting}

@@

\item \textbf{Abstract data types:} \( \ires \) supports only three abstract
data types: \( \code{Record} \) for mappings from some values to other values,
\( \code{List} \) for sequential data, and \( \code{Symbol} \) for singleton
data.  For example, ECMAScript environment records are represented as \(
\code{Record} \) from string values into addresses that represent other types
of \( \code{Record} \) for bindings.
\end{itemize}
We define syntax of \( \ires \) including 15 kinds of instructions and 26 kinds
of expressions with the notation \( \irinst \) and \( \irexpr \), respectively.
Moreover, we also formally define their operational semantics with the form \(
\irstate \vdash \irinst \Rightarrow \irstate \) for instructions, and the form
\( \irstate \vdash \irexpr \Rightarrow (\irvalue, \irstate) \) for expressions
where \( \irvalue \) denotes a value and \( \irstate \) denotes a state.
However, we omit this formalization in this paper for brevity, instead, we
include it in the companion material\footnote{\inred{???}}.

\subsection{Implementation}
We implemented {\sf Algorithm Compiler} by extending Packrat
parsing~\cite{packrat} library in Scala parser combinators. We modified the
meaning of the composition operator ( \( | \) ) to collect all the
longest matched results.  If a parser detects a step that cannot be parsed or
parsed in multiple ways, it reports the step with the parsing results.

\paragraph{Compile Rules}
\textsf{Algorithm Compiler} requires compile rules to compile given abstract
algorithms to \( \ires \) functions.  As already explained in
Section~\ref{sec:overview}, we found general patterns in writing style of
abstract algorithms.  We manually define the general compile rules to represent
such writing style with six different kinds as summarized in
Table~\ref{table:rules}.  The compile rule for statements, \( \code{Stmt} \),
generates \( \ires \) instructions.  The \( \code{Expr} \), \( \code{Cond} \),
and \( \code{Value} \) compile rules generate \( \ires \) expressions, but they
represents different contexts in ECMAScript; \( \code{Expr} \) represents a
context where any expression can appear, \( \code{Cond} \) denotes a context
where any boolean-valued expression can appear, and \( \code{Value} \)
represents a context where a fully evaluated value can appear.  The \( \code{Ty}
\) compile rule denotes type names and generates string primitives used in
object constructions.  The \( \code{Ref} \) represents references such as
identifier lookup and member accesses of objects, and it generates \( \ires \)
references.

@@

\paragraph{Global Setting}
To construct {\sf AST-\( \ires \) Translator}, global settings are also required
and it consists of \textit{ECMAScript data types} and \textit{built-in objects}.
Unlike compile rules, global settings are slightly different depending on
ECMAScript versions. In this paper, we construct global settings only for the
most recent ECMAScript version, ES10.

ECMAScript describes data types with several fields and methods used in the
specification.  Their methods are abstract algorithms but their semantics are
slightly different with other. They implicitly get their receiver objects as
arguments at callsites.  To mimic such implicit behavior, we added a special
variable \( \code{this} \) as the first parameter of a method, and passed a
receiver object at its callsite by modifying \textsf{Algorithm Compiler}.  For
example, an Environment Record type has the \textbf{DeleteBinding (N)} method.
Thus, the corresponding \( \ires \) function has two parameters, the special
parameter \( \code{this} \) and normal one \( \code{N} \) and the method call
\raisebox{-0.65ex}{\includegraphics[width=0.17\textwidth]{img/method-call.png}}
in an abstract algorithm is compiled to the \( \ires \) instruction:
\code{(DclRec.DeleteBinding DclRec N)}

In ECMAScript, built-in objects are pre-defined functions with several
built-in functions.  For example, \( \code{Array} \)
is the constructor of array objects, and its prototype
\( \code{Array.prototype} \) has built-in functions for array objects.
For instance, \( \code{[1,2,3].flat()} \) calls the
\( \code{Array.prototype.flat} \) built-in function with the array
\( \code{[1,2,3]} \).  Because built-in functions are also abstract
algorithms, each of them is automatically converted to an \( \ires \)
function.  However, the structures of built-in objects should be
manually implemented.  Thus, we implemented built-in objects in Scala
and connected their properties with the extracted \( \ires \) functions.
%
Some built-in objects that are explicitly referenced in abstract
algorithms are intrinsic objects, which have their own
aliased names summarized in Table 7 of Section 6.1.7.4
\textsf{Well-Known Intrinsic Objects}.  We extracted the alias
into \textsf{Global Setting} to utilize it during evaluation.

>>>>

\section{Algorithm Compiler}\label{sec:compiler}
\begin{figure}
  \centering
  \includegraphics[width=0.48\textwidth]{img/algo-compiler.pdf}
  \caption{Overall structure of {\sf Algorithm Compiler}}
  \label{fig:algo-compiler}
\end{figure}

@@

Moreover, {\sf Tokenizer} flattens a structured step to a single token
list to handle multi-step statements easily.  Some statements in
abstract algorithms consist of multiple steps.  For example, the
\( \code{if}\!-\!\code{then}\!-\!\code{else} \) statement often consists of
two steps: one for the \( \code{then} \)-branch and the other
for the \( \code{else} \)-branch.  To treat them as a linear structure,
we introduce three special tokens to break down structured algorithms:
\( \tend \) denotes the end of a single step, and \( \tin \) and
\( \tout \) denote the start and the end of nested steps, respectively.
For example, the following left abstract algorithm is tokenized to the right
token list.
\[
  \begin{array}{lcl}
    \code{1. A}\\
    \code{2. B} &\Longrightarrow& \code{A} \tend \code{B} \tin \code{C} \tend \tout \tend\\
    \code{\ \ a. C}\\
  \end{array}
\]

@@

\subsection{Token List Parser}
The token list parser is defined with \textit{parsing rules}. A parsing rule
is a basic parsing rule or a composition of multiple parsing rules. The composition
\code{A | B} of two parsing rules \code{A} and \code{B} parses an input using
both rules and collects the longest matched results.  If both rules fail or
match the same length of the input, the composition fails.

We provide two kinds of basic parsing rules: \textit{tag-based rules} and
\textit{content-based rules}.  A tag-based rule just checks whether the next
token has a given tag.  For example, the tag-based parser \( \code{varT} \) and
\( \code{codeT} \) check whether the next token has the tag \( \code{<var>} \)
and \( \code{<code>} \), respectively.  A content-based parser checks whether
the next token is a text token and its content passes a given condition.  For
example, the string literal \( \code{"Perform"} \) denotes a content-based
parser that checks whether the next token is a text token with the content \(
\code{Perform} \).  We also define two content-based parsers \( \code{word} \)
and \( \code{number} \) that check whether the content of the next token
consists of only alphabets or numbers, respectively.  In addition, we provide
several helper functions such as the optional rule \code{A?} and
the positive(negative) predicate \code{+A}(\code{-A}).  For instance, the helper
function \( \code{repsep(A, B)} \) generates a new parsing rule that denotes
zero or more repetition of the parsing rule \( \code{A} \) using another parsing
rule \( \code{B} \) as a separator.

@@

\subsection{Token AST Converter}
\textit{Conversion rules} describe how to generate an \( \ires \) function for a
given token AST.  Each conversion rule is defined with its corresponding parsing
rule.  For basic parsing rules, their conversion rules always return the string
values of the contents in parsed tokens.  For example, the following conversion
rules are the omitted parts in the previous example for the step 5 of
\textbf{Evaluation}:
\begin{lstlisting}[style=myScalastyle]
// statements
val Stmt = ... ^^ { case _ ~ e ~ _ => IExpr(e) }
// expressions
val Expr =
  // codes      // false literal
  ... ^^ EStr | ... ^^ { _ => EBool(false) }            |
  // variables  // additions
  ... ^^ EId  | ... ^^ { case x ~ _ ~ y => EAdd(x, y) } |
  // function calls
  ... ^^ { case x ~ _ ~ y ~ _ => ECall(x, y) }
)
\end{lstlisting}
The conversion rule of the \code{Stmt} compile rule uses only the second sub
tree and constructs an \code{IExpr} \( \ires \) instruction.  For the second
sub-tree, the conversion rule of the fifth \code{Expr} compile rule is
applied. It constructs \code{ECall} \( \ires \) expression with the string value
of the first sub-tree and the sequence of the expressions of the third sub-tree.
In this way, the step 5 of \textbf{Evaluation} is converted to the following \(
\ires \) instruction whose beautified form is the seventh line
in Figure~\ref{fig:array-literal-eval}(b).
\begin{lstlisting}[style=ires]
IExpr(ECall(EId("Set"), List(
  EId(array), EStr("length"), ECall(EId(ToUint32), List(
    EAdd(EId("padding"), EId("len")))), EBool(false))))
\end{lstlisting}

@@

\item \textbf{Abstract data types:} \( \ires \) supports only three abstract
data types: \( \code{Record} \) for mappings from values to values,
\( \code{List} \) for sequential data, and \( \code{Symbol} \) for singleton
data.  For example, ECMAScript environment records are represented as \(
\code{Record} \) from string values to addresses that represent the
bindings of the string values.
\end{itemize}
We define the syntax of \( \ires \) that has 15 kinds of instructions and 26 kinds
of expressions with the notation \( \irinst \) and \( \irexpr \), respectively.
We also formally define its operational semantics \(
\irstate \vdash \irinst \Rightarrow \irstate \) for instructions and
\( \irstate \vdash \irexpr \Rightarrow (\irvalue, \irstate) \) for expressions,
where \( \irstate \) denotes a state and \( \irvalue \) denotes a value.
For presentation brevity, we omit the formalization of \( \ires \) in this paper
and include it in a companion report\inred{\cite{report}}.

\subsection{Implementation}
We implemented {\sf Algorithm Compiler} by extending the Packrat
parsing~\cite{packrat} library in Scala parser combinators. We modified the
meaning of the composition operator ( \( | \) ) to collect all the
longest matched results.  If a parser detects a step that cannot be parsed or
is parsed in multiple ways, it reports the step with parsing results.

\paragraph{Compile Rules}
\textsf{Algorithm Compiler} requires compile rules to compile given abstract
algorithms to \( \ires \) functions.  As already explained in
Section~\ref{sec:overview}, we found common patterns in the writing style of
abstract algorithms.  We manually defined general compile rules to represent
such a writing style with six different kinds as summarized in
Table~\ref{table:rules}.  The compile rule for statements, \( \code{Stmt} \),
generates \( \ires \) instructions.  The \( \code{Expr} \), \( \code{Cond} \),
and \( \code{Value} \) compile rules generate \( \ires \) expressions, but they
represent different contexts in ECMAScript; \( \code{Expr} \) represents a
context where any expression can appear, \( \code{Cond} \) denotes a context
where any boolean-valued expression can appear, and \( \code{Value} \)
represents a context where a fully evaluated value can appear.  The \( \code{Ty}
\) compile rule denotes type names and generates string primitives used in
object constructions.  The \( \code{Ref} \) compile rule represents references such as
identifier lookup and member accesses of objects, and it generates \( \ires \)
references.

@@

\paragraph{Global Setting}
{\sf AST-\( \ires \) Translator} uses global settings consisting
of \textit{ECMAScript data types} and \textit{built-in objects}.
Unlike compile rules, global settings depend on
ECMAScript versions. In this paper, we construct global settings only for the
latest ECMAScript, ES10.

ECMAScript describes data types with some fields and methods.
While the methods are like abstract algorithms, their semantics are
slightly different from abstract algorithms. They implicitly get their receiver objects as
arguments at callsites.  To mimic such an implicit behavior, we added a special
variable \( \code{this} \) as the first parameter of each method, and passed a
receiver object at its callsite by modifying \textsf{Algorithm Compiler}.  For
example, an Environment Record type has the \textbf{DeleteBinding (N)} method.
Thus, its corresponding \( \ires \) function has two parameters, the special
parameter \( \code{this} \) and a normal parameter \( \code{N} \), and the method call
\raisebox{-0.65ex}{\includegraphics[width=0.17\textwidth]{img/method-call.png}}
in an abstract algorithm is compiled to the \( \ires \) instruction:
\code{(DclRec.DeleteBinding DclRec N)}.

In ECMAScript, built-in objects are pre-defined functions with several
built-in functions.  For example, \( \code{Array} \)
is the constructor of array objects, and its prototype
\( \code{Array.prototype} \) has built-in functions for array objects.
For instance, \( \code{[1,2,3].flat()} \) calls the
\( \code{Array.prototype.flat} \) built-in function with the array
\( \code{[1,2,3]} \).  Because built-in functions are also abstract
algorithms, each of them is automatically converted to an \( \ires \)
function.  However, the structures of built-in objects should be
manually implemented.  Thus, we implemented built-in objects in Scala
and connected their properties with the extracted \( \ires \) functions.
%
Some built-in objects that are explicitly referenced in abstract
algorithms are intrinsic objects, which have their own
aliased names summarized in Table~7 of Section~6.1.7.4
\textsf{Well-Known Intrinsic Objects}.  We extracted the alias
into \textsf{Global Setting} to utilize it during evaluation.
